Gut microbes and their genes or something else

### 2023-02-17 manuscript/tex/main.tex
clean up

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{multirow}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage[normalem]{ulem}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

@@

\hypertarget{title}{%
\subsection{Title}\label{title}}

\begin{itemize}
\item
  \begin{quote}
  Gut microbes and their genes are associated with cognition and
  neuroanatomy in children
  \end{quote}
\item
  \begin{quote}
  Gut microbes and their genes are associated with cognition performance
  and brain structure in children
  \end{quote}
\item
  \begin{quote}
  Extensive relationships between the gut microbiota and normal
  neurocognitive development
  \end{quote}
\item
  \begin{quote}
  Extensive relationships between the gut microbiota and typical
  neurocognitive development
  \end{quote}
\item
  \begin{quote}
  The gut microbial metagenome is connected to neurocognition in typical
  child development
  \end{quote}
\item
  \begin{quote}
  The microbial potential in regulating early childhood neurocognitive
  development
  \end{quote}
\end{itemize}

\hypertarget{authors}{%
\subsubsection{Authors}\label{authors}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{quote}
  Bonham 0000-0003-3200-7533
  \end{quote}
\item
  \begin{quote}
  Bottino 0000-0003-1953-1576
  \end{quote}
\item
  \begin{quote}
  McCann 0000-0002-9753-7968
  \end{quote}
\item
  \begin{quote}
  Beauchemin \{ORCID??\}
  \end{quote}
\item
  \begin{quote}
  Elizabeth Weisse \textless elizabeth.weisse2@gmail.com\textgreater{}
  \end{quote}
\item
  \begin{quote}
  \ldots{}
  \end{quote}
\item
  \begin{quote}
  Jonathan O\textquotesingle Muircheartaigh
  \textless{}\href{mailto:jonathanom@kcl.ac.uk}{\uline{jonathanom@kcl.ac.uk}}\textgreater{}
  \end{quote}
\item
  \begin{quote}
  Huttenhower 0000-0002-1110-0096
  \end{quote}
\item
  \begin{quote}
  Muriel Bruchhage
  \textless{}\href{mailto:mmk.bruchhage@gmail.com}{\uline{mmk.bruchhage@gmail.com}}\textgreater{}
  0000-0001-9637-0951
  \end{quote}
\item
  \begin{quote}
  D'Sa 0000-0002-3476-8419
  \end{quote}
\item
  \begin{quote}
  Deoni 0000-0002-0633-6328
  \end{quote}
\item
  \begin{quote}
  Klepac-Ceraj 0000-0001-5387-5706
  \end{quote}
\end{enumerate}

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

The gastrointestinal tract, its resident microorganisms and the central
nervous system are connected by biochemical signaling, also known as
"microbiome-gut-brain-axis." Both the human brain and the gut microbiome
have critical developmental windows in the first three years of life,
raising the possibility that their development is co-occurring and
likely co-dependent. Emerging evidence implicates gut microorganisms and
microbiota composition in cognitive outcomes and neurodevelopmental
disorders (e.g., autism and anxiety), but the influence of gut microbial
metabolism on typical neurodevelopment has not been explored in detail.
We investigated the relationship of the microbiome with the neuroanatomy
and cognitive function of 361 healthy children, demonstrating that
differences in gut microbial taxa and gene functions are associated with
overall cognitive function as well as with differences in the size of
multiple brain regions. Using a combination of multivariate linear
models and machine learning (ML) models, we showed that many species,
including \emph{Gordonibacter pamelae} and \emph{Blautia wexlerae}, were
significantly associated with higher cognitive function, while some
species such as \emph{Ruminococcus gnavus} were more commonly found in
children with low cognitive scores regardless of age or maternal
education. Microbial genes for enzymes involved in the metabolism of
neuroactive compounds, particularly short-chain fatty acids such as
glutamate and propionate, were also associated with cognitive function.
In addition, ML models were able to use microbial taxa to predict the
volume of brain regions, with particular taxa often dominating the
feature importance metric for individual brain regions, such as \emph{B.
wexlerae} being the most dominant for the parahippocampal region or
GABA-producing \emph{Bacteroidetes ovatus} for left accumbens. These
findings provide potential biomarkers of neurocognition and may lead to
the future development of targets for early detection and early
intervention.

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

The gut and the brain are intimately linked. Signals from the brain
reach the gut through the autonomic nervous system and the endocrine
system, and the gut can communicate with the brain through the vagus
nerve and through endocrine and immune (cytokine) signaling molecules
\href{https://www.zotero.org/google-docs/?i3SRPn}{(Cerdó et al., 2019;
Pronovost and Hsiao, 2019; Sharon et al., 2016; Tognini, 2017)}. In
addition, the products of microbial metabolism generated in the gut can
influence the brain, both indirectly by stimulating the enteric nervous
and immune systems, as well as directly through molecules that enter
circulation and cross the blood-brain barrier. Causal links between the
gut microbiome and neural development, particularly atypical
development, are increasingly being identified
\href{https://www.zotero.org/google-docs/?UsrFKB}{(Spichak et al.,
2021)}. Both human epidemiology and animal models point to the effects
of gut microbes on the development of autism spectrum disorder
\href{https://www.zotero.org/google-docs/?Y8l5A0}{(Laue et al., 2020;
Wan et al., 2021)} and particular microbial taxa have been associated
with depression
\href{https://www.zotero.org/google-docs/?mbaFE3}{(Mayneris-Perxachs et
al., 2022; Valles-Colomer et al., 2019)} and Alzheimer's disease
\href{https://www.zotero.org/google-docs/?0BEGUn}{(Fung et al., 2017;
Kim et al., 2021)}. But information about this ``microbiome-gut-brain
axis'' in normal neurocognitive development remains lacking,
particularly early in life.

The first years of life are critical developmental windows for both the
microbiome and the brain
\href{https://www.zotero.org/google-docs/?XTqhlm}{(Laue et al., 2022)}.
Fetal development \emph{in utero} is believed to be mostly sterile, but
is rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food sources (breastmilk or formula),
and other environmental sources such as antibiotics
\href{https://www.zotero.org/google-docs/?X6N0f4}{(Bäckhed et al., 2015;
Bokulich et al., 2016; Dominguez-Bello et al., 2010; Louwies et al.,
2020)}. The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\href{https://www.zotero.org/google-docs/?VG43uh}{(Koenig et al.,
2011)}. Many of these microbes have specialized metabolic capabilities
for digesting human breast milk, such as \emph{Bifidobacterium infantis}
and \emph{Bacteroides fragilis}
\href{https://www.zotero.org/google-docs/?1dKzKj}{(Sela et al., 2008;
Tso et al., 2021)}. Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation, with most taxa
of the infant microbiome being replaced by taxa more reminiscent of
adult microbiomes
\href{https://www.zotero.org/google-docs/?V3lY3p}{(Bäckhed et al.,
2015)}. Many prior studies have focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
neural synaptogenesis, investigation across this solid-food boundary is
incredibly important
\href{https://www.zotero.org/google-docs/?d29bv8}{(Tau and Peterson,
2010)}.

A child's brain undergoes remarkable anatomical, microstructural,
organizational, and functional changes in the first years of life. By
age 5, a child's brain has reached \textgreater85\% of its adult size,
has achieved near-adult levels of myelination, and the pattern of axonal
connections has been established
\href{https://www.zotero.org/google-docs/?1jUZOO}{(Silbereis et al.,
2016)}. Much of this development occurs in discrete windows or
``sensitive-'' or ``critical periods'' (CPs)
\href{https://www.zotero.org/google-docs/?w4Pcze}{(Knudsen, 2004)} when
neural plasticity is particularly high, and particular modes of learning
and skill development are preferred. The timing of these sensitive
periods is driven in part by genetics, but can also be affected by the
environment, including the gut microbiome
\href{https://www.zotero.org/google-docs/?KTm9lQ}{(Cowan et al., 2020)}.
In fact, emerging evidence suggests that the timing and duration of CPs
may be driven in part by cues from the developing gut microbiome
\href{https://www.zotero.org/google-docs/?mAD1ne}{(Callaghan, 2020)}. As
such, understanding the normal spectrum of healthy microbiome
development and how it relates to normal neurocognitive development may
provide opportunities for identifying atypical development earlier and
offer opportunities for intervention.

To begin to address this need, we investigated the gut microbiome and
neurocognitive development of children from infancy through 10 years of
age in an accelerated longitudinal study design. Gut microbial
communities were assessed using shotgun metagenomic sequencing, enabling
profiling at both the taxonomic and gene-functional level, and
neurocognitive development was measured using expert-assessment of
cognitive function and neuroimaging using magnetic resonance imaging
(MRI). Using a combination of classical statistical analysis and machine
learning, we showed that the development of the gut microbiome, the
brain, and children's cognitive abilities are intimately linked, with
both microbial taxa and gene functions able to predict cognitive
performance and brain structure.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{cohort-overview-and-summary-data}{%
\subsubsection{Cohort overview and summary
data}\label{cohort-overview-and-summary-data}}

We investigated the co-development of the brain and the microbiome in
early childhood in a cohort of typically-developing children in their
first years of life using a variety of orthogonal microbial and
neurocognitive assessments, including shotgun metagenomic sequencing,
cognitive and behavioral assessments, and neuroimaging. In all, 361
children from RESONANCE, an accelerated longitudinal cohort of early
development \href{https://paperpile.com/c/dPbU4e/klxU}{(Forrest,
Blackwell, and Camargo 2018)} between the ages of 82 days and 10 years
old (median age 2.21 years, Supplementary Figure 1) were included in
this study (Table 1, Figure 1A). To measure cognitive function, we used
age-appropriate assessments that can be normalized to a common, IQ-like
scale (Figure 1B), including the Mullen scale of early learning (MSEL)
for children from birth to 3 years of age
\href{https://www.zotero.org/google-docs/?90hfYr}{(Mullen and others,
1995)}, Wechsler preschool and primary scale of intelligence (WPPSI) for
4--5-year-olds
\href{https://www.zotero.org/google-docs/?wCfx5h}{(Wechsler, 2012)}, and
the Wechsler intelligence scale for children (WISC) for children 6 years
and up \href{https://www.zotero.org/google-docs/?AcNN3S}{(Wechsler,
1949)}.

As expected, the greatest differences in microbial taxa were driven by
age (Figure 1B-C, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-\emph{Prevotella copri} (Supplementary Figure 2). Overall variation
in gut microbial genes and also in brain volume profiles was similarly
driven largely by subject age (Figure 1C-F).

\includegraphics[width=6.45833in,height=3.87221in]{media/image3.png}

\textbf{Figure 1: Cohort overview and summary data.}

(A) Study design and overview. Stool samples, cognitive assessments and
neuroimaging were collected from participants in an accelerated
longitudinal design. (B) Cognitive function scores are assessed by
different instruments, but may be normalized to a common IQ-like scale.
(C-D) Principal coordinates analysis using Bray-Curtis dissimilarity on
taxonomic profiles demonstrates high beta diversity, with much of the
first axis of variation explained by increasing age and alpha diversity.
Differences in gene function profiles (E) and neuroimaging (PCA based on
the Euclidean distance of brain region volumes) (F) are likewise
dominated by changes as children age. (G) Permutation analysis of
variance (PERMANOVA) of taxonomic profiles, functional profiles
(annotated as UniRef90s) and neuroimaging, against the metadata of
interest. Variations in taxonomic and functional profiles explain a
modest but significant percent of the variation in cognitive development
in children over 18 months of age. (H) Mantel testing of different
microbial feature matrices, shows overlapping but distinct patterns of
variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
used as cut-offs in some following models.

\textbf{Table1: Subject demographics}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1845}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2620}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1845}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1845}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1845}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Group}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Subgroup}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{All}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Under 6 mo}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Over 18 mo}
\end{minipage} \\
\begin{minipage}[b]{\linewidth}\raggedright
N subjects
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
361
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
74
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
257
\end{minipage} \\
\multirow{3}{*}{\begin{minipage}[b]{\linewidth}\raggedright
Samples
\end{minipage}} & \begin{minipage}[b]{\linewidth}\raggedright
1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
262 (72.6\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
74 (100.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
203 (79.0\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
72 (19.9\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
0 (0.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
46 (17.9\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
\textgreater2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
27 (7.5\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
0 (0.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
8 (3.1\%)
\end{minipage} \\
\multirow{3}{*}{\begin{minipage}[b]{\linewidth}\raggedright
Age (months)
\end{minipage}} & \begin{minipage}[b]{\linewidth}\raggedright
Min
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
2.8
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
2.8
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
18.0
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Max
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
119.3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
6.0
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
119.3
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Median
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
25.8
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
3.7
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
45.9
\end{minipage} \\
\multirow{2}{*}{\begin{minipage}[b]{\linewidth}\raggedright
Sex
\end{minipage}} & \begin{minipage}[b]{\linewidth}\raggedright
F
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
163 (45.2\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
37 (50.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
114 (44.4\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
M
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
198 (54.8\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
37 (50.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
143 (55.6\%)
\end{minipage} \\
\multirow{6}{*}{\begin{minipage}[b]{\linewidth}\raggedright
Race
\end{minipage}} & \begin{minipage}[b]{\linewidth}\raggedright
White
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
252 (69.8\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
42 (56.8\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
191 (74.3\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Black
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
40 (11.1\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
16 (21.6\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
21 (8.2\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Asian
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (0.3\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
0 (0.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (0.4\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Mixed
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
7 (1.9\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (1.4\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
6 (2.3\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Other
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
48 (13.3\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
12 (16.2\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
29 (11.3\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Unknown/Declined
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
2 (0.6\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (1.4\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
2 (0.8\%)
\end{minipage} \\
\multirow{6}{*}{\begin{minipage}[b]{\linewidth}\raggedright
Maternal Ed.
\end{minipage}} & \begin{minipage}[b]{\linewidth}\raggedright
Junior high school
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (0.3\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
0 (0.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
1 (0.4\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Some high school
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
6 (1.7\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
3 (4.1\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
4 (1.6\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
High school grad
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
37 (10.2\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
13 (17.6\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
16 (6.2\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Some college
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
95 (26.3\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
25 (33.8\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
57 (22.2\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
College grad
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
90 (24.9\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
17 (23.0\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
69 (26.8\%)
\end{minipage} \\
& \begin{minipage}[b]{\linewidth}\raggedright
Grad/professional school
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
123 (34.1\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
15 (20.3\%)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
104 (40.5\%)
\end{minipage} \\
\midrule()
\endhead
\bottomrule()
\end{longtable}

Several studies have demonstrated links between specific taxa and
measures of anxiety and depression
\href{https://www.zotero.org/google-docs/?X5rfKj}{(Mayneris-Perxachs et
al., 2022; Needham et al., 2022)}, cognitive flexibility in adults
\href{https://www.zotero.org/google-docs/?RGIf8m}{(Magnusson et al.,
2015)}, and atypical neural development
\href{https://www.zotero.org/google-docs/?hwPZKh}{(Liu et al., 2019; Wan
et al., 2021)}. We, therefore, set out to identify whether specific taxa
or gene functions were linked with normal cognitive development in
children. In order to assess whether variations in gut microbial taxa,
their genes, or their metabolism are linked with neurocognitive
development, we tested whether the beta diversity of microbial taxa and
gene functions, as well as variation in brain development as assessed by
neuroimaging, were associated with these contemporaneous measures of
cognitive function using permutation analysis of variance (PERMANOVA).
Due to the large ecological shift in the microbiome that occurs upon the
introduction of solid food, and a relatively wide range of ages when
infants are transitioned to solid food, we considered the pre-transition
(less than 6 months old) and post-transition (over 18 months old)
microbiomes separately. Even still, age was a major driver of variation
in gut microbiomes in children over 18 months old (Figure 1G). We also
found that overall variation in microbial species in children over 18
months old was associated with a small but significant variation in
cognitive function score (Figure 1G, R\textsuperscript{2} = 0.0124, q
\textless{} 0.001), as was variation in microbial gene functions
(R\textsuperscript{2} = 0.0119, q \textless{} 0.001). Variation in
microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity, and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and UniRef90s
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

Consistent with prior studies, different microbial measurement types
captured overlapping variation, with species profiles and gene function
profiles (annotated with clusters of 90\% similarity
\href{https://www.zotero.org/google-docs/?b9rEJS}{(Suzek et al.,
2007)}), both generated from DNA sequencing data, being tightly coupled
(Figure 1H, p \textless{} 0.001). Interestingly, other functional
groupings (Enzyme commission level4 - ECs
\href{https://www.zotero.org/google-docs/?fs8LO7}{(Bairoch, 2000)}, and
KEGG orthologues - KOs
\href{https://www.zotero.org/google-docs/?Xx5duE}{(Kanehisa et al.,
2004)}) overlapped only slightly with taxonomic profiles in both age
cohorts, despite being derived from UniRef90 labels. In children over 18
months, some variation (15.9\%, p \textless{} 0.01) in neuroimaging
overlapped with microbial measures, though this may be due to the
residual variation due to age in both measures.

\hypertarget{microbial-species-and-neuroactive-genes-are-associated-with-cognitive-performance}{%
\subsubsection{Microbial species and neuroactive genes are associated
with cognitive
performance}\label{microbial-species-and-neuroactive-genes-are-associated-with-cognitive-performance}}

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression
\href{https://www.zotero.org/google-docs/?4qS6UF}{(Mallick et al.,
2021)} to the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N = 92 for 0--120 months, N
= 46 for 0--6 months, N = 97 for 18--120 months). Only \emph{Blautia
wexlerae} was significantly associated (q value = 0.14, β = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B). \emph{B. wexlerae} was
previously shown to be depleted in children with diabetes
\href{https://www.zotero.org/google-docs/?6mU2nI}{(Benítez-Páez et al.,
2020)}, and that oral administration of \emph{B. wexlerae} partially
ameliorated weight gain and inflammation from a high-fat diet in a mouse
model of T2D \href{https://www.zotero.org/google-docs/?y83hpS}{(Hosomi
et al., 2022; Liu et al., 2021)}. In children over 18 months of age,
several microbial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}, which produces the
neuroprotective metabolite urolithin
\href{https://www.zotero.org/google-docs/?HGtTEY}{(Gong et al., 2022;
Selma et al., 2014)}, \emph{Asaccharobacter celatus} and
\emph{Adelcreutzia equolifaciens}, which produce phytoestrogen-derived
equol \href{https://www.zotero.org/google-docs/?HCKZqm}{(Maruo et al.,
2008; Thawornkuno et al., 2009)}, and the SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii} \href{https://www.zotero.org/google-docs/?3eyheK}{(Ghosh et
al., 2020; Lopez-Siles et al., 2017; Mukherjee et al., 2020)} (Figure
2B).

\includegraphics[width=6.5in,height=4.83333in]{media/image4.png}

\textbf{Figure 2 - Taxa and gene functional groups are associated with
cognitive function.}

(A) Volcano plot of multivariable linear model results showing the
relationship between individual taxa and cognitive function score in
children over 18 months of age, controlling for age, maternal education,
and sequencing depth. Taxa that were significant after FDR correction (q
\textless{} 0.2) are colored red. (B) For taxa that were significantly
associated with cognitive function, heatmaps of prevalence, mean
relative abundance, and correlation with cognitive function in different
age groups. (C-F) Enrichment plots for selected neuroactive gene sets
used in feature set enrichment analysis (FSEA). (G) summary of FSEA
results across all samples (left) as well as in the under 6 months
(middle) and over 18 months (right) subsets, colored based on the
significance of the association. Markers indicate the individual
correlation of genes within a gene set, and vertical bars represent the
median correlation of that gene set.

>>>>

%%%%%%%%%%TO FORMAT A PREPRINT FOR A PCI%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[a4paper]{article}

\input{preamble.tex}

\title{Gut microbes and their genes or something else}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%  SET THE LIST OF AUTHORS WITH CORRESPONDING AFFILIATIONS  use \& before last author %%%%%%%%%%%%%%%%%%%%
\author{Kevin S. Bonham\textsuperscript{1}, \space
Guilherme Fahur Bottino\textsuperscript{1}, 
Firstname Lastname\textsuperscript{2}, \&
Vanja Klepac-Ceraj\textsuperscript{1}
%\fourthauthor \textsuperscript{i}
%etc...
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  SET THE ABSTRACT %%%%%%%%%%%%%%%%%

@@

\begin{abstract}
    The gastrointestinal tract, its resident microorganisms and the central
    nervous system are connected by biochemical signaling, also known as
    "microbiome-gut-brain-axis." Both the human brain and the gut microbiome
    have critical developmental windows in the first three years of life,
    raising the possibility that their development is co-occurring and
    likely co-dependent. Emerging evidence implicates gut microorganisms and
    microbiota composition in cognitive outcomes and neurodevelopmental
    disorders (e.g., autism and anxiety), but the influence of gut microbial
    metabolism on typical neurodevelopment has not been explored in detail.
    We investigated the relationship of the microbiome with the neuroanatomy
    and cognitive function of 361 healthy children, demonstrating that
    differences in gut microbial taxa and gene functions are associated with
    overall cognitive function as well as with differences in the size of
    multiple brain regions. Using a combination of multivariate linear
    models and machine learning (ML) models, we showed that many species,
    including \emph{Gordonibacter pamelae} and \emph{Blautia wexlerae}, were
    significantly associated with higher cognitive function, while some
    species such as \emph{Ruminococcus gnavus} were more commonly found in
    children with low cognitive scores regardless of age or maternal
    education. Microbial genes for enzymes involved in the metabolism of
    neuroactive compounds, particularly short-chain fatty acids such as
    glutamate and propionate, were also associated with cognitive function.
    In addition, ML models were able to use microbial taxa to predict the
    volume of brain regions, with particular taxa often dominating the
    feature importance metric for individual brain regions, such as \emph{B.
    wexlerae} being the most dominant for the parahippocampal region or
    GABA-producing \emph{Bacteroidetes ovatus} for left accumbens. These
    findings provide potential biomarkers of neurocognition and may lead to
    the future development of targets for early detection and early
    intervention.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%% Text of the preprint %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% use \emph{} for italics and \parencite{} to cite a reference and \ref{XX} to cite the \label{XX}
% use \\ and a blank line to mark the end of paragraph and starting of a new one

\section*{Introduction}

\input{sections/intro}

\section*{Results}

\input{sections/results}

\section*{Discussion}

\input{sections/discussion}

\section*{Material and methods}

\input{sections/methods}
